---
title: "Spatial Join of Urban Features to Base Network"
author: "Christine"
date: "2026-03-03"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

## 1. Introduction & Methodology

This document outlines the spatial data processing workflow for assigning various urban facilities (points/small polygons) to a street centerline network (`basenetwork`). 

Because the real-world coordinates of facilities rarely perfectly overlap with street centerlines, we employ two distinct spatial join strategies based on the physical distribution of the features:

* **Edge-based Features (Nearest Neighbor):** For facilities distributed along road segments (e.g., street poles, transit stops, traffic calming devices), we snap them to the nearest street centerline.
* **Node-based Features (Buffer Intersect):** For facilities located at intersections (e.g., red light cameras, intersection controls), we create a **30-foot buffer**. 
    * *Justification for 30ft:* Standard urban street right-of-way (ROW) is typically 40-60 feet. A 30-foot radius from a street corner (where cameras/signs are physically installed) creates a 60-foot diameter coverage. This perfectly bridges the geometric gap between the physical corner and the topological center of the intersection, ensuring the node feature is accurately assigned to all 3-4 intersecting street segments without bleeding into adjacent blocks.

---

## 2. Environment Setup & Authentication

```{r setup, message=FALSE, warning=FALSE}
# --- Environment Setup & Box Authentication ---
library(tidyverse)
library(tidygeocoder)
library(sf)
library(mapview) 
library(janitor) 
library(boxr)

# You only need to authenticate once per R session.
# Ensure BOX_CLIENT_ID and SECRET are configured in your .Renviron
box_auth()
```

## 3. Custom Functions

We define a robust helper function to handle downloading and reading various spatial formats from Box, automatically standardizing them to **EPSG:2272 (NAD83 / Pennsylvania South)**.

```{r define_functions}
# --- Smart Box Spatial Data Reader ---
read_box_sf <- function(file_id, crs = 2272) {
  # 1. Get temporary directory path
  temp_dir <- tempdir()
  
  # 2. Download file and get exact local path
  dl_info <- box_dl(file_id, local_dir = temp_dir, overwrite = TRUE)
  file_path_exact <- as.character(dl_info)
  
  # 3. Determine file type and read accordingly
  if (grepl("\\.rds$", file_path_exact, ignore.case = TRUE)) {
    # If it's an .rds file, use readRDS
    cat("Detected .rds file, reading with readRDS...\n")
    sf_data <- readRDS(file_path_exact)
    
    # Ensure the object is an 'sf' object before transforming CRS
    if (inherits(sf_data, "sf")) {
      sf_data <- st_transform(sf_data, crs)
    } else {
      warning("Warning: This .rds file does not contain sf spatial data!")
    }
    
  } else {
    # If it's a standard spatial file (.geojson, .shp), use st_read
    cat("Detected standard spatial file, reading with st_read...\n")
    sf_data <- st_read(file_path_exact, quiet = TRUE) %>% 
      st_transform(crs)
  }
  
  return(sf_data)
}
```

## 4. Data Loading & Cleaning

```{r load_data, results='hide'}
# --- Read data and clean column names (converts to lowercase with underscores) ---
basenetwork <- read_box_sf(2151757279199) %>% clean_names()

traffic_calming <- read_box_sf(2123213862368) %>% clean_names()
intersection_ctrl <- read_box_sf(2123208654111) %>% clean_names()
street_poles <- read_box_sf(2129860050884) %>% clean_names()
transit_stops <- read_box_sf(2130613247190) %>% clean_names()
fixed_camera <- read_box_sf(2144281173122) %>% clean_names()

# Set the unique identifier column for the base network
unique_id_col <- "seg_id"
```

## 5. Strategy A: Edge-Based Features (Nearest Neighbor)

Applicable to features distributed along road segments (street poles, transit stops, traffic calming). We use `st_nearest_feature` to match each point to the closest street centerline.

```{r strategy_a_nearest}
# --- Conflict-proof nearest neighbor counting function ---
count_nearest_points <- function(points_sf, lines_sf, id_col, new_count_name) {
  
  # [Crucial Step]: Only keep the geometry of the point data!
  # This entirely prevents .x and .y renaming issues caused by identical column names
  points_only_geom <- points_sf %>% select(geometry)
  
  # Keep only the unique ID column for the line data
  lines_only_id <- lines_sf %>% select(all_of(id_col))
  
  # Spatial nearest neighbor join
  points_joined <- st_join(points_only_geom, lines_only_id, join = st_nearest_feature)
  
  # Aggregate and count
  summary_df <- points_joined %>%
    st_drop_geometry() %>%
    filter(!is.na(!!sym(id_col))) %>% # Filter out anomalies where ID is NA
    group_by(!!sym(id_col)) %>%
    summarise(!!new_count_name := n(), .groups = "drop")
  
  return(summary_df)
}

# --- Execute calculation ---
poles_count <- count_nearest_points(street_poles, basenetwork, unique_id_col, "count_poles")
bus_count <- count_nearest_points(transit_stops, basenetwork, unique_id_col, "count_transit")
calming_count <- count_nearest_points(traffic_calming, basenetwork, unique_id_col, "count_calming")
```

## 6. Strategy B: Node-Based Features (Buffer Intersection & Visualization)

This strategy handles features located at intersections, such as red light cameras and stop signs. Since these facilities govern all approaching street segments, we apply a **30-foot buffer** to capture all intersecting centerlines rather than just the single closest one.

```{r strategy_b_buffer_and_viz, message=FALSE, warning=FALSE}
# --- 1. Define Function for Intersection Buffer Join ---
count_intersection_points <- function(points_sf, lines_sf, id_col, new_count_name, buffer_ft = 30) {
  # Generate buffers around intersection features
  points_buffered <- st_buffer(points_sf, dist = buffer_ft)
  
  # Perform spatial intersection (joining lines to buffers)
  lines_joined <- st_join(lines_sf, points_buffered, join = st_intersects)
  
  # Aggregate counts by street segment ID
  summary_df <- lines_joined %>%
    st_drop_geometry() %>%
    # Filter out segments that did not intersect any buffer
    filter(!is.na(st_drop_geometry(points_sf)[1, 1])) %>% 
    group_by(!!sym(id_col)) %>%
    summarise(!!new_count_name := n(), .groups = "drop")
  
  return(summary_df)
}

# --- 2. Execute Spatial Joins for Node-Based Datasets ---
intersect_count <- count_intersection_points(intersection_ctrl, basenetwork, unique_id_col, "count_intersection_ctrl")
camera_count <- count_intersection_points(fixed_camera, basenetwork, unique_id_col, "count_camera")

# --- 3. Interactive Visualization of the Buffer Logic ---
# Extract a sample of 10 points for demonstration to maintain performance
sample_points <- intersection_ctrl %>% head(10)

# Generate 30ft buffers for the sample
sample_buffers <- st_buffer(sample_points, dist = 30)

# Identify street centerlines that intersect the sample buffers
intersecting_lines <- basenetwork %>% 
  st_filter(sample_buffers, .predicate = st_intersects)

# Create an interactive map using mapview
# Layering: Lines (Bottom) -> Buffers (Middle) -> Points (Top)
mapview(intersecting_lines, color = "royalblue", lwd = 4, layer.name = "Intersecting Streets") +
  mapview(sample_buffers, col.regions = "red", alpha.regions = 0.5, layer.name = "30ft Buffers") +
  mapview(sample_points, color = "black", col.regions = "black", cex = 1, layer.name = "Facility Nodes")
```

## 7. Final Assembly

Merge all the newly created count features back into the original street network. Replace `NA` values with `0` for road segments that do not have these facilities.

```{r final_join}
# --- Join all features back to the base network ---
basenetwork_final <- basenetwork %>%
  left_join(poles_count, by = unique_id_col) %>%
  left_join(bus_count, by = unique_id_col) %>%
  left_join(calming_count, by = unique_id_col) %>%
  left_join(intersect_count, by = unique_id_col) %>%
  left_join(camera_count, by = unique_id_col) %>%
  mutate(across(starts_with("count_"), ~replace_na(.x, 0)))

# Check the structure of the final data
glimpse(basenetwork_final)
```