---
title: "Philadelphia Crash Analysis (2020-2024)"
subtitle: "MUSA Practicum x OTIS"
author: "Kavana Raju"
date: "`r Sys.Date()`"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    theme: cosmo
---

## Research Question

Do roads designed for peak capacity contribute to catastrophic crash outcomes during off-peak hours?

**Hypothesis**: Roads designed for peak traffic volumes have excess capacity during off-peak periods, enabling higher speeds and more severe crashes.

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(sf)
library(scales)
library(boxr)
library(patchwork)

# Suppress progress bars globally
options(
  scipen = 999,
  boxr.progress = FALSE,
  readr.show_progress = FALSE,
  dplyr.show_progress = FALSE
)

# Authenticate with Box (suppress messages)
suppressMessages(box_auth())

# Plot themes
plotTheme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  plot.subtitle = element_text(size = 10),
  plot.caption = element_text(size = 8, hjust = 0, lineheight = 1.1),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 11, face = "bold"),
  panel.background = element_blank(),
  panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
  panel.grid.minor = element_blank(),
  axis.ticks = element_blank(),
  legend.position = "right"
)

mapTheme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  plot.subtitle = element_text(size = 10),
  plot.caption = element_text(size = 8),
  axis.line = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank(),
  axis.title = element_blank(),
  panel.background = element_blank(),
  panel.border = element_blank(),
  panel.grid.major = element_line(colour = 'transparent'),
  panel.grid.minor = element_blank(),
  legend.position = "right",
  plot.margin = margin(1, 1, 1, 1, 'cm')
)

severity_colors <- c(
  "Fatal" = "#d62728",
  "Suspected Serious Injury" = "#ff7f0e", 
  "Suspected Minor Injury" = "#ffbb78",
  "Possible Injury" = "#c5b0d5",
  "PDO" = "#c7c7c7"
)

# Helper function to read spatial data from Box
read_box_sf <- function(file_id, crs = 2272) {
  temp_dir <- tempdir()
  dl_file_path <- box_dl(file_id, local_dir = temp_dir, overwrite = TRUE)
  sf_data <- st_read(dl_file_path, quiet = TRUE) %>% st_transform(crs)
  return(sf_data)
}
```

---

## Part 1: Load and Harmonize Crash Data

```{r load-data, message=FALSE}
# 2020 data
crash_2020_id <- "2129578929490"
flags_2020_id <- "2129571870046"
roadway_2020_id <- "2129574335626"

# 2021 data
crash_2021_id <- "2129584283678"
flags_2021_id <- "2129572071618"
roadway_2021_id <- "2129575576179"

# 2022 data
crash_2022_id <- "2129572753523"
flags_2022_id <- "2129573139646"
roadway_2022_id <- "2129573617700"

# 2023 data
crash_2023_id <- "2129590204560"
flags_2023_id <- "2129570178906"
roadway_2023_id <- "2129571655183"

# 2024 data
crash_2024_id <- "2129573871742"
flags_2024_id <- "2129575475146"
roadway_2024_id <- "2129573394142"

# Load all years from Box
crash_2020 <- box_read_csv(crash_2020_id)
flags_2020 <- box_read_csv(flags_2020_id)
roadway_2020 <- box_read_csv(roadway_2020_id)

crash_2021 <- box_read_csv(crash_2021_id)
flags_2021 <- box_read_csv(flags_2021_id)
roadway_2021 <- box_read_csv(roadway_2021_id)

crash_2022 <- box_read_csv(crash_2022_id)
flags_2022 <- box_read_csv(flags_2022_id)
roadway_2022 <- box_read_csv(roadway_2022_id)

crash_2023 <- box_read_csv(crash_2023_id)
flags_2023 <- box_read_csv(flags_2023_id)
roadway_2023 <- box_read_csv(roadway_2023_id)

crash_2024 <- box_read_csv(crash_2024_id)
flags_2024 <- box_read_csv(flags_2024_id)
roadway_2024 <- box_read_csv(roadway_2024_id)

# Harmonize data types (WEATHER1 and TCD_FUNC_CD changed across years)
crash_years <- list(crash_2020, crash_2021, crash_2022, crash_2023, crash_2024)
crash_years_harmonized <- crash_years %>%
  map(~ mutate(.x, 
               WEATHER1 = as.character(WEATHER1),
               TCD_FUNC_CD = as.character(TCD_FUNC_CD)))

# Combine all years
all_crashes <- bind_rows(crash_years_harmonized)
all_flags <- bind_rows(flags_2020, flags_2021, flags_2022, flags_2023, flags_2024)
all_roadway <- bind_rows(roadway_2020, roadway_2021, roadway_2022, roadway_2023, roadway_2024)

# Filter roadway for primary road (RDWY_SEQ_NUM = 3)
all_roadway_primary <- all_roadway %>%
  filter(RDWY_SEQ_NUM == 3) %>%
  select(CRN, ROUTE, OFFSET, STREET_NAME, SPEED_LIMIT, LANE_COUNT, ROAD_OWNER)

# Join tables
crash_master <- all_crashes %>%
  left_join(all_flags, by = "CRN") %>%
  left_join(all_roadway_primary, by = "CRN")

# Filter to Philadelphia (COUNTY = "67")
philly_crashes <- crash_master %>%
  filter(COUNTY == "67")

cat("✓ Loaded and filtered to Philadelphia\n")
cat("  Total crashes:", format(nrow(philly_crashes), big.mark = ","), "\n")
cat("  By year:\n")
print(table(philly_crashes$CRASH_YEAR))
```

---

## Part 2: Feature Engineering

```{r feature-engineering}
philly_crashes <- philly_crashes %>%
  mutate(
    # Hour of crash (extract from TIME_OF_DAY, fallback to DISPATCH_TM)
    hour_raw = case_when(
      !is.na(TIME_OF_DAY) ~ str_sub(as.character(TIME_OF_DAY), 1, 2),
      !is.na(DISPATCH_TM) ~ str_sub(as.character(DISPATCH_TM), 1, 2),
      TRUE ~ NA_character_
    ),
    HOUR_OF_DAY_CLEAN = as.integer(hour_raw),
    HOUR_OF_DAY_CLEAN = if_else(HOUR_OF_DAY_CLEAN == 99, NA_integer_, HOUR_OF_DAY_CLEAN),
    
    # KSI flag (Killed or Suspected Serious Injury)
    is_ksi = if_else(FATAL_OR_SUSP_SERIOUS_INJ == "1", 1, 0, missing = 0),
    
    # Time periods (peak: 7-10 AM, 4-7 PM)
    time_period = case_when(
      is.na(HOUR_OF_DAY_CLEAN) ~ "Unknown",
      HOUR_OF_DAY_CLEAN >= 7 & HOUR_OF_DAY_CLEAN < 10 ~ "AM Peak (7-10)",
      HOUR_OF_DAY_CLEAN >= 16 & HOUR_OF_DAY_CLEAN < 19 ~ "PM Peak (16-19)",
      HOUR_OF_DAY_CLEAN >= 10 & HOUR_OF_DAY_CLEAN < 16 ~ "Midday (10-16)",
      HOUR_OF_DAY_CLEAN >= 19 & HOUR_OF_DAY_CLEAN < 22 ~ "Evening (19-22)",
      HOUR_OF_DAY_CLEAN >= 22 | HOUR_OF_DAY_CLEAN < 7 ~ "Night/Late (22-7)",
      TRUE ~ "Other"
    ),
    
    is_peak = if_else(
      time_period %in% c("AM Peak (7-10)", "PM Peak (16-19)"), 
      1, 0, missing = 0
    ),
    
    # Day of week
    day_name = case_when(
      DAY_OF_WEEK == 1 ~ "Sunday",
      DAY_OF_WEEK == 2 ~ "Monday",
      DAY_OF_WEEK == 3 ~ "Tuesday",
      DAY_OF_WEEK == 4 ~ "Wednesday",
      DAY_OF_WEEK == 5 ~ "Thursday",
      DAY_OF_WEEK == 6 ~ "Friday",
      DAY_OF_WEEK == 7 ~ "Saturday",
      TRUE ~ NA_character_
    ),
    is_weekend = if_else(DAY_OF_WEEK %in% c(1, 7), 1, 0, missing = 0),
    
    # Severity labels
    severity_label = case_when(
      MAX_SEVERITY_LEVEL == "1" ~ "Fatal",
      MAX_SEVERITY_LEVEL == "2" ~ "Suspected Serious Injury",
      MAX_SEVERITY_LEVEL == "3" ~ "Suspected Minor Injury",
      MAX_SEVERITY_LEVEL == "4" ~ "Possible Injury",
      MAX_SEVERITY_LEVEL == "0" ~ "PDO",
      TRUE ~ "Unknown"
    ),
    
    # Convert flags to integer
    PEDESTRIAN = as.integer(PEDESTRIAN),
    BICYCLE = as.integer(BICYCLE),
    VULNERABLE_ROAD_USER = as.integer(VULNERABLE_ROAD_USER),
    SPEEDING_RELATED = as.integer(SPEEDING_RELATED),
    ILLUMINATION_DARK = as.integer(ILLUMINATION_DARK),
    
    # Road design variables
    speed_limit = as.integer(SPEED_LIMIT),
    lane_count = case_when(
      LANE_COUNT == "99" ~ NA_integer_,
      TRUE ~ as.integer(LANE_COUNT)
    ),
    
    # Coordinate validity
    has_valid_coords = !is.na(DEC_LATITUDE) & !is.na(DEC_LONGITUDE),
    
    # Clean street name for matching
    crash_street_clean = str_to_upper(str_trim(STREET_NAME))
  ) %>%
  
  # Set factor levels
  mutate(
    time_period = factor(
      time_period,
      levels = c("AM Peak (7-10)", "Midday (10-16)", "PM Peak (16-19)",
                 "Evening (19-22)", "Night/Late (22-7)", "Unknown")
    ),
    
    severity_label = factor(
      severity_label,
      levels = c("Fatal", "Suspected Serious Injury", "Suspected Minor Injury",
                 "Possible Injury", "PDO", "Unknown")
    )
  )

cat("✓ Feature engineering complete\n")
```

### Summary Statistics

```{r summary-stats}
cat("Dataset Summary:\n")
cat("  Total crashes:", format(nrow(philly_crashes), big.mark = ","), "\n")
cat("  KSI crashes:", format(sum(philly_crashes$is_ksi), big.mark = ","), 
    "(", round(mean(philly_crashes$is_ksi) * 100, 1), "%)\n")
cat("  Crashes with time data:", format(sum(!is.na(philly_crashes$HOUR_OF_DAY_CLEAN)), big.mark = ","),
    "(", round(mean(!is.na(philly_crashes$HOUR_OF_DAY_CLEAN)) * 100, 1), "%)\n")
cat("  Crashes with coordinates:", format(sum(philly_crashes$has_valid_coords), big.mark = ","),
    "(", round(mean(philly_crashes$has_valid_coords) * 100, 1), "%)\n")
cat("  Crashes with speed limit:", format(sum(!is.na(philly_crashes$speed_limit)), big.mark = ","),
    "(", round(mean(!is.na(philly_crashes$speed_limit)) * 100, 1), "%)\n")
cat("  Crashes with lane count:", format(sum(!is.na(philly_crashes$lane_count)), big.mark = ","),
    "(", round(mean(!is.na(philly_crashes$lane_count)) * 100, 1), "%)\n")

```
---

## Data Quality Assessment

### Missing Data Analysis
```{r missing-data-analysis}

missing_summary <- data.frame(
  Variable = c("HOUR_OF_DAY_CLEAN", "speed_limit", "lane_count", 
               "DEC_LATITUDE", "DEC_LONGITUDE", "PEDESTRIAN", 
               "BICYCLE", "SPEEDING_RELATED", "ALCOHOL_RELATED"),
  Missing_Count = c(
    sum(is.na(philly_crashes$HOUR_OF_DAY_CLEAN)),
    sum(is.na(philly_crashes$speed_limit)),
    sum(is.na(philly_crashes$lane_count)),
    sum(is.na(philly_crashes$DEC_LATITUDE)),
    sum(is.na(philly_crashes$DEC_LONGITUDE)),
    sum(is.na(philly_crashes$PEDESTRIAN)),
    sum(is.na(philly_crashes$BICYCLE)),
    sum(is.na(philly_crashes$SPEEDING_RELATED)),
    sum(is.na(philly_crashes$ALCOHOL_RELATED))
  )
)

missing_summary$Missing_Percent <- round(missing_summary$Missing_Count / nrow(philly_crashes) * 100, 1)

knitr::kable(missing_summary, 
             col.names = c("Variable", "Missing Count", "Missing %"),
             caption = "Missing Data Summary")
```

### Feature Distributions
```{r feature-distributions, fig.width=12, fig.height=8}
# Speed limit distribution
p1 <- ggplot(philly_crashes %>% filter(!is.na(speed_limit)), 
             aes(x = as.factor(speed_limit))) +
  geom_bar(fill = "#156082") +
  labs(title = "Crashes by Posted Speed Limit",
       x = "Speed Limit (mph)", y = "Count") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 0))

# Lane count distribution
p2 <- ggplot(philly_crashes %>% filter(!is.na(lane_count)), 
             aes(x = as.factor(lane_count))) +
  geom_bar(fill = "#ff9500") +
  labs(title = "Crashes by Lane Count",
       x = "Number of Lanes", y = "Count") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 0))

# Hour distribution
p3 <- ggplot(philly_crashes %>% filter(!is.na(HOUR_OF_DAY_CLEAN)), 
             aes(x = HOUR_OF_DAY_CLEAN)) +
  geom_bar(fill = "#028090") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Crashes by Hour of Day",
       x = "Hour (0-23)", y = "Count") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 0))

# Day of week distribution
p4 <- ggplot(philly_crashes %>% filter(!is.na(day_name)), 
             aes(x = factor(day_name, levels = c("Sunday", "Monday", "Tuesday", 
                                 "Wednesday", "Thursday", "Friday", "Saturday")))) +
  geom_bar(fill = "#d62728") +
  labs(title = "Crashes by Day of Week",
       x = NULL, y = "Count") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45))

(p1 + p2) / (p3 + p4)
```
---

## Part 3: Road Network Linkage

**Two-Tier Approach**: (1) Street name matching (high confidence), (2) Spatial snap within 100 feet (fallback)

```{r road-linkage, message=FALSE, warning=FALSE}
# Load Street Centerline shapefile
street_centerline_id <- "2137970615238"

street_roads <- read_box_sf(street_centerline_id, crs = 2272) %>%
  select(seg_id, st_name, st_type, class, responsibl, geometry) %>%
  mutate(
    full_street_name = str_trim(paste(st_name, st_type)),
    full_street_name_clean = str_to_upper(full_street_name)
  )

street_lookup <- street_roads %>% 
  st_drop_geometry() %>%
  distinct(full_street_name_clean, seg_id, st_name, st_type)

# Tier 1: Street name matching
# Strategy: For each crash, keep only the FIRST matching segment to avoid memory explosion
crashes_street_matched <- philly_crashes %>%
  filter(!is.na(crash_street_clean)) %>%
  inner_join(
    street_lookup %>% 
      group_by(full_street_name_clean) %>% 
      slice(1) %>%  # Keep only first segment per street name
      ungroup(),
    by = c("crash_street_clean" = "full_street_name_clean")
  ) %>%
  mutate(road_source = "Street_Name", match_confidence = "High")

# Verify no duplicates
cat("Tier 1 deduplication check:", sum(duplicated(crashes_street_matched$CRN)), "duplicates remaining\n")

# Tier 2: Spatial snap (for unmatched crashes with coordinates)
crashes_needing_snap <- philly_crashes %>%
  filter(!(CRN %in% crashes_street_matched$CRN), has_valid_coords) %>%
  st_as_sf(coords = c("DEC_LONGITUDE", "DEC_LATITUDE"), crs = 4326) %>%
  st_transform(crs = 2272)

crashes_snapped <- crashes_needing_snap %>%
  mutate(
    nearest_road_id = st_nearest_feature(geometry, street_roads),
    snap_distance = as.numeric(st_distance(geometry, street_roads[nearest_road_id,], by_element = TRUE))
  ) %>%
  filter(snap_distance <= 100) %>%
  mutate(
    road_source = "Spatial_Snap",
    match_confidence = case_when(
      snap_distance <= 50 ~ "High",
      snap_distance <= 75 ~ "Medium",
      TRUE ~ "Low"
    )
  )

# Combine and add to philly_crashes
tier1_df <- crashes_street_matched %>%
  select(CRN, seg_id, road_source, match_confidence)

tier2_df <- crashes_snapped %>%
  mutate(seg_id = street_roads$seg_id[nearest_road_id]) %>%
  st_drop_geometry() %>%
  select(CRN, seg_id, road_source, match_confidence, snap_distance)

# Combine tier1 and tier2 - ensure no duplicates
combined_linkage <- bind_rows(tier1_df, tier2_df) %>%
  distinct(CRN, .keep_all = TRUE)

cat("\nFinal deduplication check:", sum(duplicated(combined_linkage$CRN)), "duplicates\n")

philly_crashes <- philly_crashes %>%
  left_join(combined_linkage, by = "CRN", relationship = "one-to-one") %>%
  mutate(has_road_link = !is.na(road_source))

cat("Road Linkage Summary:\n")
cat("  Tier 1 (street name):", format(nrow(tier1_df), big.mark = ","), "\n")
cat("  Tier 2 (spatial snap):", format(nrow(tier2_df), big.mark = ","), "\n")
cat("  Total linked:", format(sum(philly_crashes$has_road_link), big.mark = ","),
    "(", round(mean(philly_crashes$has_road_link) * 100, 1), "%)\n")
```

---

## Part 4: Temporal Patterns

```{r hourly-patterns, fig.width=10, fig.height=6}
crashes_timed <- philly_crashes %>%
  filter(!is.na(HOUR_OF_DAY_CLEAN))

hourly_summary <- crashes_timed %>%
  group_by(HOUR_OF_DAY_CLEAN) %>%
  summarise(
    total_crashes = n(),
    ksi_crashes = sum(is_ksi),
    ksi_rate = ksi_crashes / total_crashes * 100,
    .groups = "drop"
  )

ggplot(hourly_summary, aes(x = HOUR_OF_DAY_CLEAN)) +
  geom_col(aes(y = total_crashes), fill = "#156082", alpha = 0.7) +
  geom_line(aes(y = ksi_rate * 50), color = "#d62728", linewidth = 1.2) +
  geom_point(aes(y = ksi_rate * 50), color = "#d62728", size = 3) +
  scale_y_continuous(
    name = "Total Crashes",
    sec.axis = sec_axis(~./50, name = "KSI Rate (%)")
  ) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Hourly Crash Patterns in Philadelphia (2020-2024)",
    subtitle = "Total crashes (bars) and KSI rate (line)",
    x = "Hour of Day",
    caption = "Peak hours: 7-10 AM, 4-7 PM | KSI = Killed or Suspected Serious Injury"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 0))
```

---

## Part 5: Peak vs Off-Peak Analysis

```{r peak-offpeak-analysis}
peak_comparison <- philly_crashes %>%
  filter(!is.na(HOUR_OF_DAY_CLEAN)) %>%
  mutate(period_type = if_else(is_peak == 1, "Peak Hours", "Off-Peak Hours")) %>%
  group_by(period_type) %>%
  summarise(
    total_crashes = n(),
    ksi_crashes = sum(is_ksi),
    ksi_rate = ksi_crashes / total_crashes * 100,
    speeding_crashes = sum(SPEEDING_RELATED, na.rm = TRUE),
    speeding_rate = speeding_crashes / total_crashes * 100,
    ped_crashes = sum(PEDESTRIAN, na.rm = TRUE),
    ped_rate = ped_crashes / total_crashes * 100,
    .groups = "drop"
  )

knitr::kable(peak_comparison, digits = 1, caption = "Peak vs Off-Peak Crash Characteristics")

# Statistical test (chi-squared)
ksi_test <- prop.test(
  x = c(sum(philly_crashes$is_ksi[philly_crashes$is_peak == 0], na.rm = TRUE),
        sum(philly_crashes$is_ksi[philly_crashes$is_peak == 1], na.rm = TRUE)),
  n = c(sum(philly_crashes$is_peak == 0, na.rm = TRUE),
        sum(philly_crashes$is_peak == 1, na.rm = TRUE))
)

cat("\nChi-squared test for KSI rate difference:\n")
cat("  p-value:", format.pval(ksi_test$p.value), "\n")
cat("  Statistically significant:", if_else(ksi_test$p.value < 0.05, "Yes", "No"), "\n")
```

### Visualization

```{r peak-viz, fig.width=8, fig.height=5}
ggplot(peak_comparison, aes(x = period_type, y = ksi_rate, fill = period_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(ksi_rate, 1), "%")), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Peak Hours" = "#156082", "Off-Peak Hours" = "#ff9500")) +
  labs(
    title = "KSI Rate: Peak vs Off-Peak Hours",
    y = "KSI Rate (%)",
    x = NULL
  ) +
  plotTheme +
  theme(legend.position = "none", axis.text.x = element_text(angle = 0))
```

---

## Part 6: Speed Limit Interaction

```{r speed-interaction}
speed_peak_summary <- philly_crashes %>%
  filter(!is.na(speed_limit), !is.na(HOUR_OF_DAY_CLEAN)) %>%
  mutate(period_type = if_else(is_peak == 1, "Peak", "Off-Peak")) %>%
  group_by(speed_limit, period_type) %>%
  summarise(
    total_crashes = n(),
    ksi_crashes = sum(is_ksi),
    ksi_rate = ksi_crashes / total_crashes * 100,
    .groups = "drop"
  ) %>%
  filter(total_crashes >= 20)

ggplot(speed_peak_summary, aes(x = as.factor(speed_limit), y = ksi_rate, fill = period_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Peak" = "#156082", "Off-Peak" = "#ff9500")) +
  labs(
    title = "KSI Rate by Speed Limit and Time Period",
    subtitle = "Only roads with 20+ crashes shown",
    x = "Posted Speed Limit (mph)",
    y = "KSI Rate (%)",
    fill = "Period"
  ) +
  plotTheme
```

---

## Part 7: Spatial Visualization

```{r prepare-spatial-data}
# Load Philadelphia boundary from Box
city_limits_id <- "2187037802115"
philly_boundary <- read_box_sf(city_limits_id, crs = 2272)

# Create crash points
crash_points <- philly_crashes %>%
  filter(has_valid_coords) %>%
  st_as_sf(coords = c("DEC_LONGITUDE", "DEC_LATITUDE"), crs = 4326) %>%
  st_transform(crs = 2272)
```

### Map 1: Crash Points by Severity

```{r map-all-crashes, fig.width=12, fig.height=10}
ggplot() +
  geom_sf(data = philly_boundary, fill = "white", color = "gray50") +
  geom_sf(data = crash_points, aes(color = severity_label), alpha = 0.6, size = 0.5) +
  scale_color_manual(values = severity_colors, name = "Severity") +
  labs(
    title = "All Crash Locations in Philadelphia (2020-2024)",
    subtitle = paste0(format(nrow(crash_points), big.mark = ","), " crashes with valid coordinates"),
    caption = "Source: PennDOT Crash Database"
  ) +
  mapTheme
```

### Map 2: KSI Crashes Only

```{r spatial-maps, fig.width=12, fig.height=10}
ksi_points <- crash_points %>% filter(is_ksi == 1)

ggplot() +
  geom_sf(data = philly_boundary, fill = "white", color = "gray50") +
  geom_sf(data = ksi_points, aes(color = severity_label), alpha = 0.6, size = 0.5) +
  scale_color_manual(values = severity_colors, name = "Severity") +
  labs(
    title = "KSI Crashes in Philadelphia (2020-2024)",
    subtitle = paste0(format(nrow(ksi_points), big.mark = ","), " crashes"),
    caption = "Source: PennDOT Crash Database"
  ) +
  mapTheme
```

---

## Part 8: Segment-Level Analysis and Export

```{r segment-aggregation}
# Just filter for high-confidence matches
crashes_for_segments <- philly_crashes %>%
  filter(match_confidence == "High", !is.na(seg_id))

# Comprehensive segment aggregation
segment_attributes_full <- crashes_for_segments %>%
  group_by(seg_id) %>%
  summarise(
    # Total counts
    total_crashes = n(),
    crashes_2020 = sum(CRASH_YEAR == 2020, na.rm = TRUE),
    crashes_2021 = sum(CRASH_YEAR == 2021, na.rm = TRUE),
    crashes_2022 = sum(CRASH_YEAR == 2022, na.rm = TRUE),
    crashes_2023 = sum(CRASH_YEAR == 2023, na.rm = TRUE),
    crashes_2024 = sum(CRASH_YEAR == 2024, na.rm = TRUE),
    
    # Severity
    ksi_crashes = sum(is_ksi == 1, na.rm = TRUE),
    fatal_crashes = sum(MAX_SEVERITY_LEVEL == "1", na.rm = TRUE),
    serious_inj = sum(MAX_SEVERITY_LEVEL == "2", na.rm = TRUE),
    minor_inj = sum(MAX_SEVERITY_LEVEL == "3", na.rm = TRUE),
    poss_inj = sum(MAX_SEVERITY_LEVEL == "4", na.rm = TRUE),
    pdo_crashes = sum(MAX_SEVERITY_LEVEL == "0", na.rm = TRUE),
    
    # VRU
    ped_crashes = sum(PEDESTRIAN == 1, na.rm = TRUE),
    bike_crashes = sum(BICYCLE == 1, na.rm = TRUE),
    vru_crashes = sum(VULNERABLE_ROAD_USER == 1, na.rm = TRUE),
    
    # Characteristics
    speeding = sum(SPEEDING_RELATED == 1, na.rm = TRUE),
    alcohol = sum(ALCOHOL_RELATED == 1, na.rm = TRUE),
    dark = sum(ILLUMINATION_DARK == 1, na.rm = TRUE),
    
    # Temporal
    peak_crash = sum(is_peak == 1, na.rm = TRUE),
    offpk_crash = sum(is_peak == 0 & !is.na(is_peak), na.rm = TRUE),
    am_peak = sum(time_period == "AM Peak (7-10)", na.rm = TRUE),
    pm_peak = sum(time_period == "PM Peak (16-19)", na.rm = TRUE),
    midday = sum(time_period == "Midday (10-16)", na.rm = TRUE),
    evening = sum(time_period == "Evening (19-22)", na.rm = TRUE),
    night = sum(time_period == "Night/Late (22-7)", na.rm = TRUE),
    weekend = sum(is_weekend == 1, na.rm = TRUE),
    weekday = sum(is_weekend == 0, na.rm = TRUE),

    # Hourly crash counts (0-23)
    h00 = sum(HOUR_OF_DAY_CLEAN == 0, na.rm = TRUE),
    h01 = sum(HOUR_OF_DAY_CLEAN == 1, na.rm = TRUE),
    h02 = sum(HOUR_OF_DAY_CLEAN == 2, na.rm = TRUE),
    h03 = sum(HOUR_OF_DAY_CLEAN == 3, na.rm = TRUE),
    h04 = sum(HOUR_OF_DAY_CLEAN == 4, na.rm = TRUE),
    h05 = sum(HOUR_OF_DAY_CLEAN == 5, na.rm = TRUE),
    h06 = sum(HOUR_OF_DAY_CLEAN == 6, na.rm = TRUE),
    h07 = sum(HOUR_OF_DAY_CLEAN == 7, na.rm = TRUE),
    h08 = sum(HOUR_OF_DAY_CLEAN == 8, na.rm = TRUE),
    h09 = sum(HOUR_OF_DAY_CLEAN == 9, na.rm = TRUE),
    h10 = sum(HOUR_OF_DAY_CLEAN == 10, na.rm = TRUE),
    h11 = sum(HOUR_OF_DAY_CLEAN == 11, na.rm = TRUE),
    h12 = sum(HOUR_OF_DAY_CLEAN == 12, na.rm = TRUE),
    h13 = sum(HOUR_OF_DAY_CLEAN == 13, na.rm = TRUE),
    h14 = sum(HOUR_OF_DAY_CLEAN == 14, na.rm = TRUE),
    h15 = sum(HOUR_OF_DAY_CLEAN == 15, na.rm = TRUE),
    h16 = sum(HOUR_OF_DAY_CLEAN == 16, na.rm = TRUE),
    h17 = sum(HOUR_OF_DAY_CLEAN == 17, na.rm = TRUE),
    h18 = sum(HOUR_OF_DAY_CLEAN == 18, na.rm = TRUE),
    h19 = sum(HOUR_OF_DAY_CLEAN == 19, na.rm = TRUE),
    h20 = sum(HOUR_OF_DAY_CLEAN == 20, na.rm = TRUE),
    h21 = sum(HOUR_OF_DAY_CLEAN == 21, na.rm = TRUE),
    h22 = sum(HOUR_OF_DAY_CLEAN == 22, na.rm = TRUE),
    h23 = sum(HOUR_OF_DAY_CLEAN == 23, na.rm = TRUE),
    time_unknown = sum(is.na(HOUR_OF_DAY_CLEAN)),

    # KSI by time period
    ksi_peak = sum(is_ksi == 1 & is_peak == 1, na.rm = TRUE),
    ksi_offpk = sum(is_ksi == 1 & is_peak == 0, na.rm = TRUE),
    ksi_am = sum(is_ksi == 1 & time_period == "AM Peak (7-10)", na.rm = TRUE),
    ksi_pm = sum(is_ksi == 1 & time_period == "PM Peak (16-19)", na.rm = TRUE),
    ksi_midday = sum(is_ksi == 1 & time_period == "Midday (10-16)", na.rm = TRUE),
    ksi_eve = sum(is_ksi == 1 & time_period == "Evening (19-22)", na.rm = TRUE),
    ksi_night = sum(is_ksi == 1 & time_period == "Night/Late (22-7)", na.rm = TRUE),
    ksi_wknd = sum(is_ksi == 1 & is_weekend == 1, na.rm = TRUE),
    ksi_wkdy = sum(is_ksi == 1 & is_weekend == 0, na.rm = TRUE),
    
    # Road design
    spd_lim_md = as.numeric(names(sort(table(speed_limit), decreasing = TRUE))[1]),
    lane_ct_md = median(lane_count, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Rates
    ksi_rate = if_else(total_crashes > 0, ksi_crashes / total_crashes * 100, 0),
    fatal_rate = if_else(total_crashes > 0, fatal_crashes / total_crashes * 100, 0),
    ped_rate = if_else(total_crashes > 0, ped_crashes / total_crashes * 100, 0),
    bike_rate = if_else(total_crashes > 0, bike_crashes / total_crashes * 100, 0),
    vru_rate = if_else(total_crashes > 0, vru_crashes / total_crashes * 100, 0),
    spd_rate = if_else(total_crashes > 0, speeding / total_crashes * 100, 0),
    alc_rate = if_else(total_crashes > 0, alcohol / total_crashes * 100, 0),
    dark_rate = if_else(total_crashes > 0, dark / total_crashes * 100, 0),
    
    # Temporal shares
    offpk_share = if_else(
      (peak_crash + offpk_crash) > 0,
      offpk_crash / (peak_crash + offpk_crash) * 100,
      NA_real_
    ),
    wknd_share = if_else(total_crashes > 0, weekend / total_crashes * 100, 0),
    night_share = if_else(total_crashes > 0, night / total_crashes * 100, 0),

    # KSI rates by time period
    ksi_rate_peak = if_else(peak_crash > 0, ksi_peak / peak_crash * 100, NA_real_),
    ksi_rate_offpk = if_else(offpk_crash > 0, ksi_offpk / offpk_crash * 100, NA_real_),
    ksi_rate_night = if_else(night > 0, ksi_night / night * 100, NA_real_),
    ksi_rate_wknd = if_else(weekend > 0, ksi_wknd / weekend * 100, NA_real_),
    
    # Density
    crash_py = total_crashes / 5,
    ksi_py = ksi_crashes / 5
  )

cat("✓ Segment aggregation complete\n")
cat("  Segments with crashes:", format(nrow(segment_attributes_full), big.mark = ","), "\n")
```

### Segment-Level Statistics

```{r segment-statistics}
cat("\nSegment-level summary:\n")
cat("Total segments with crashes:", format(nrow(segment_attributes_full), big.mark = ","), "\n")
cat("Segments with 1+ crashes:", sum(segment_attributes_full$total_crashes > 0), "\n")
cat("Segments with 5+ crashes:", sum(segment_attributes_full$total_crashes >= 5), "\n")
cat("Segments with 10+ crashes:", sum(segment_attributes_full$total_crashes >= 10), "\n")

# Top 10 segments by crash count
cat("\nTop 10 segments by total crashes:\n")
segment_attributes_full %>%
  arrange(desc(total_crashes)) %>%
  head(10) %>%
  select(seg_id, total_crashes, ksi_crashes, ksi_rate) %>%
  knitr::kable(digits = 1)
```

### Join Crash Statistics to Road Geometry

```{r join-segment-stats-to-roads}
# Join crash statistics to road geometry
roads_with_crashes <- street_roads %>%
  left_join(segment_attributes_full, by = "seg_id") %>%
  mutate(
    # Replace NA with 0 for segments with no crashes
    total_crashes = replace_na(total_crashes, 0),
    ksi_crashes = replace_na(ksi_crashes, 0),
    ksi_rate = replace_na(ksi_rate, 0),
    ped_crashes = replace_na(ped_crashes, 0),
    bike_crashes = replace_na(bike_crashes, 0),
    speeding = replace_na(speeding, 0),
    peak_crash = replace_na(peak_crash, 0),
    offpk_crash = replace_na(offpk_crash, 0)
  )

cat("Road segments with crash data joined:", 
    format(nrow(roads_with_crashes), big.mark = ","), "\n")
```

### Map 3: Roads Colored by Crash Count

```{r map-roads-by-crash-count, fig.width=12, fig.height=10}
# Filter to segments with at least 1 crash for visualization
roads_with_crashes_filtered <- roads_with_crashes %>%
  filter(total_crashes > 0)

ggplot() +
  geom_sf(data = street_roads, color = "gray90", linewidth = 0.2) +
  geom_sf(data = roads_with_crashes_filtered, 
          aes(color = total_crashes, linewidth = total_crashes)) +
  scale_color_viridis_c(
    name = "Total Crashes\n(2020-2024)",
    option = "plasma",
    trans = "log10",
    breaks = c(1, 5, 10, 25, 50, 100)
  ) +
  scale_linewidth_continuous(
    name = "Total Crashes\n(2020-2024)",
    range = c(0.3, 2),
    trans = "log10",
    breaks = c(1, 5, 10, 25, 50, 100),
    guide = "none"
  ) +
  labs(
    title = "Philadelphia Road Segments by Crash Frequency (2020-2024)",
    subtitle = "Darker/thicker lines = more crashes",
    caption = "Source: PennDOT Crash Database | Only segments with 1+ crashes shown"
  ) +
  mapTheme +
  theme(legend.position = "right")
```

### Map 4: High Injury Network (Roads by KSI Count)

```{r map-roads-by-ksi-count, fig.width=12, fig.height=10}
# Filter to segments with at least 1 KSI crash
roads_with_ksi <- roads_with_crashes %>%
  filter(ksi_crashes > 0)

ggplot() +
  geom_sf(data = street_roads, color = "gray90", linewidth = 0.2) +
  geom_sf(data = roads_with_ksi, 
          aes(color = ksi_crashes, linewidth = ksi_crashes)) +
  scale_color_gradient(
    name = "KSI Crashes\n(2020-2024)",
    low = "#ff9500",
    high = "#d62728",
    trans = "log10",
    breaks = c(1, 2, 5, 10, 20)
  ) +
  scale_linewidth_continuous(
    name = "KSI Crashes\n(2020-2024)",
    range = c(0.5, 3),
    trans = "log10",
    breaks = c(1, 2, 5, 10, 20),
    guide = "none"
  ) +
  labs(
    title = "Philadelphia High Injury Network (2020-2024)",
    subtitle = "Road segments by KSI crash frequency",
    caption = "Source: PennDOT Crash Database | Only segments with 1+ KSI crashes shown"
  ) +
  mapTheme +
  theme(legend.position = "right")
```

### Map 5: Peak vs Off-Peak Crash Distribution by Segment

```{r map-peak-offpeak-ratio, fig.width=12, fig.height=10}
# Calculate peak/off-peak ratio for segments with 5+ crashes
roads_peak_offpeak <- roads_with_crashes %>%
  filter(total_crashes >= 5) %>%
  mutate(
    offpeak_ratio = offpk_crash / (peak_crash + offpk_crash),
    dominant_period = case_when(
      offpeak_ratio > 0.6 ~ "Off-Peak Dominant",
      offpeak_ratio < 0.4 ~ "Peak Dominant",
      TRUE ~ "Balanced"
    )
  )

ggplot() +
  geom_sf(data = street_roads, color = "gray90", linewidth = 0.2) +
  geom_sf(data = roads_peak_offpeak, 
          aes(color = offpeak_ratio), 
          linewidth = 1) +
  scale_color_gradient2(
    name = "Off-Peak\nCrash Share",
    low = "#156082",      # Blue = peak dominant
    mid = "gray70",       # Gray = balanced
    high = "#ff9500",     # Orange = off-peak dominant
    midpoint = 0.5,
    labels = percent
  ) +
  labs(
    title = "Peak vs Off-Peak Crash Distribution by Road Segment",
    subtitle = "Only segments with 5+ crashes shown",
    caption = "Blue = More peak crashes | Orange = More off-peak crashes"
  ) +
  mapTheme +
  theme(legend.position = "right")
```

### Identify High-Risk Segments

```{r high-risk-segments}
# Define high-risk segments (top 10% by crash count OR 3+ KSI crashes)
crash_threshold <- quantile(segment_attributes_full$total_crashes, 0.90, na.rm = TRUE)

high_risk_segments <- segment_attributes_full %>%
  filter(
    total_crashes >= crash_threshold | ksi_crashes >= 3
  ) %>%
  arrange(desc(ksi_crashes), desc(total_crashes))

cat("High-risk segments identified:", nrow(high_risk_segments), "\n\n")
cat("Top 20 highest-risk segments:\n")
high_risk_segments %>%
  head(20) %>%
  select(seg_id, total_crashes, ksi_crashes, ksi_rate, 
         ped_crashes, bike_crashes, speeding) %>%
  knitr::kable(digits = 1)
```

### Join to Street Centerlines and Export

```{r export-final, message=FALSE, warning=FALSE}
### Join to Street Centerlines and Export

street_centerlines_final <- street_roads %>%
  left_join(segment_attributes_full, by = "seg_id") %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Create output directory
if (!dir.exists("output")) {
  dir.create("output")
}

# Export shapefile
st_write(
  street_centerlines_final, 
  "output/street_centerlines_with_crashes.shp",
  delete_dsn = TRUE, quiet = TRUE
)

# Export GeoJSON
st_write(
  street_centerlines_final,
  "output/street_centerlines_with_crashes.geojson",
  delete_dsn = TRUE, quiet = TRUE
)

# Export CSV
street_centerlines_final %>%
  st_drop_geometry() %>%
  write_csv("output/street_segment_crash_attributes.csv")

# Export high-crash segments
high_crash_segments <- street_centerlines_final %>%
  filter(total_crashes >= 5)

st_write(
  high_crash_segments,
  "output/high_crash_segments.shp",
  delete_dsn = TRUE, quiet = TRUE
)

cat("✓ Export complete\n")
cat("  Files created:\n")
cat("    - output/street_centerlines_with_crashes.shp\n")
cat("    - output/street_centerlines_with_crashes.geojson\n")
cat("    - output/street_segment_crash_attributes.csv\n")
cat("    - output/high_crash_segments.shp\n")
```

### Final Summary

```{r final-summary}
summary_stats <- street_centerlines_final %>%
  st_drop_geometry() %>%
  summarise(
    total_segments = n(),
    segments_with_crashes = sum(total_crashes > 0),
    segments_5plus = sum(total_crashes >= 5),
    segments_10plus = sum(total_crashes >= 10),
    total_crashes = sum(total_crashes),
    total_ksi = sum(ksi_crashes),
    avg_crashes = round(mean(total_crashes), 2),
    median_crashes = median(total_crashes),
    max_crashes = max(total_crashes)
  )

knitr::kable(summary_stats, caption = "Final Dataset Summary")

cat("\nTop 20 highest-crash segments:\n")
street_centerlines_final %>%
  st_drop_geometry() %>%
  arrange(desc(total_crashes)) %>%
  select(seg_id, full_street_name, total_crashes, ksi_crashes, ksi_rate,
         ped_crashes, bike_crashes, peak_crash, offpk_crash) %>%
  head(20) %>%
  knitr::kable(digits = 1)

cat("\n✓ Processing complete - ready for modeling\n")
```

### Model-Ready Dataset Documentation
```{r model-documentation}
cat("\n=== MODEL-READY DATASET DOCUMENTATION ===\n\n")

cat("Dataset: street_centerlines_with_crashes.shp\n")
cat("Unit of Analysis: Road segment (seg_id)\n")
cat("Temporal Coverage: 2020-2024 (5 years)\n")
cat("Spatial Coverage: Philadelphia County\n\n")

cat("RECOMMENDED FILTERING FOR MODELING:\n")
cat("  Include only segments with total_crashes >= 5\n")
cat("  This provides stable rate estimates\n")
cat("  Segments meeting criteria:", sum(street_centerlines_final$total_crashes >= 5), "\n\n")

cat("TARGET VARIABLES (examples):\n")
cat("  - ksi_rate: % of crashes that are KSI (continuous, 0-100)\n")
cat("  - offpk_share: % of crashes during off-peak (continuous, 0-100)\n")
cat("  - total_crashes: Crash count (Poisson/Negative Binomial)\n")
cat("  - ksi_crashes: KSI count (Poisson/Negative Binomial)\n\n")

cat("PREDICTOR VARIABLES:\n")
cat("  Road Design: spd_lim_md, lane_ct_md, class, responsibl\n")
cat("  Crash Patterns: ped_rate, bike_rate, vru_rate, spd_rate, alc_rate, dark_rate\n")
cat("  Temporal: peak_crash, offpk_crash, wknd_share, am_peak, pm_peak, midday, evening, night\n")
cat("  Severity: fatal_rate, ksi_rate, serious_inj, minor_inj\n")
cat("  Counts: total_crashes, crashes_2020-2024, weekend, weekday\n\n")

cat("FEATURE ENGINEERING NOTES:\n")
cat("  - All rates are percentages (0-100)\n")
cat("  - Median values used for spd_lim_md and lane_ct_md (mode for speed limit)\n")
cat("  - crash_py and ksi_py are annualized (total / 5 years)\n")
cat("  - Segments with no crashes have all values set to 0\n\n")
```

---

## Validation: Crash Count Verification

```{r validation-check}
cat("\n=== FINAL VALIDATION CHECK ===\n\n")

cat("Original crash data:\n")
cat("  Total crashes loaded:", format(nrow(philly_crashes), big.mark = ","), "\n")
cat("  Crashes with road link:", format(sum(philly_crashes$has_road_link, na.rm = TRUE), big.mark = ","), "\n")
cat("  High-confidence links:", format(sum(philly_crashes$match_confidence == "High", na.rm = TRUE), big.mark = ","), "\n\n")

cat("Segment aggregations:\n")
cat("  Total segments with crashes:", format(nrow(segment_attributes_full), big.mark = ","), "\n")
cat("  Sum of all segment crashes:", format(sum(segment_attributes_full$total_crashes), big.mark = ","), "\n\n")

cat("Expected vs Actual:\n")
high_conf_crashes <- sum(philly_crashes$match_confidence == "High", na.rm = TRUE)
segment_total_crashes <- sum(segment_attributes_full$total_crashes)

cat("  High-confidence crashes:", format(high_conf_crashes, big.mark = ","), "\n")
cat("  Aggregated segment crashes:", format(segment_total_crashes, big.mark = ","), "\n")

if (abs(high_conf_crashes - segment_total_crashes) < 10) {
  cat("  ✓ VALIDATION PASSED: Counts match!\n")
} else {
  diff <- segment_total_crashes - high_conf_crashes
  cat("  ⚠ WARNING: Mismatch of", format(diff, big.mark = ","), "crashes\n")
  cat("  This suggests possible duplication or aggregation issues\n")
}
```
